D3064R0: How to Avoid OOTA Without Really Trying

### 2024-03-18 ootavacuous.tex
Record copies sent on March 18th
Olivier and Gonzalez overheard Jade and I discussing it and wanted
copies.  ;-)

Signed-off-by: Paul E. McKenney <paulmck@kernel.org>
Alan Stern March 17 feedback
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

This is not restricted to booleans, as can be seen in
Listing~\ref{lst:Unused Inventing Non-Boolean Atomic Loads}.
This example is more straightforward in that it does not make quite so
creative use of the multiple values loaded.

@@

@@

% @@@ reviewer to do.
% Luc Maranget: Memory models.
% Greg Kroah-Hartman and Linus Torvalds: FYI, trouble being caused.
% Christoph Hellwig: Why slow on BPF memory model.
% Miguel Ojeda: LKMM and Rust.
% Alice Ryhl: LKMM and Rust.
% Ralf Jung: Rust.
% Catalin Marinas: ARM and formal methods (qspinlock).
% Segher Boessenkool:  GCC.
% Dan Lustig: NVIDIA memory ordering.
% Andrea Parri: Memory models, RISC-V hardware.
% Jonas Oberhauser: Memory models.
% Hernan Ponce de Leon: Memory models.
% Palmer Dabbelt: RISC-V hardware.
% Ali Sezgin: P0422R0.
% Tony Tye: P0422R0.
% Other memory-model maintainers.
% Derek Williams: PowerPC hardware.
% Hans Boehm: Skeptic, memory models, OOTA.  Suggestions for others?
% Olivier Giroux: Skeptic, memory models, NVIDIA.
% Luke Geeson: Whatever...  Review.
% Ori Lahav: Prior OOTA.
% Viktor Vafeiadis:  Prior OOTA.
% Derek Dreyer: Prior OOTA.
% Brian Demsky: Prior OOTA.

% Done as of December 18, 2023:
% David Goldblatt
% Boqun Feng, Uladzislau Rezki, Neeraj Upadhyay, Joel Fernandes, and
%	Frederic Weisbecker (RCU proteges)
% Michael Wong
% Maged Michael
% John Wickerson: Thank you.  OK for Peter Sewell, Mark Batty, etc.
% Akira Yokosawa: Professional courtesy, perfbook memory models.
% Dan Kelley, Alexei Starovoitov, Mykola Lysenko: FYI.  OOTA dropped
% Alan Jeffrey: Fixed-point insight P0422R0.  Reached out via LinkedIn.
% Jade Alglave: C++ herd model?  Memory models and variety of hardware.
% Alan Stern: Memory models, mathematical logic, and variety of hardware.
% Martin Uecker (offered)
% Peter Sewell
% Mark Batty
% Richard Gristenthwaite: ARM hardware.  Posed HW speculation question.
% Peter Zijlstra.
% Mark Rutland: ARM memory ordering.
% Will Deacon: ARM hardware, C++, and OOTA.

>>>>

This issue is not restricted to booleans.
For example, if given the code fragment shown in
Figure~\ref{lst:OOTA Cycle}
on
page~\pageref{lst:OOTA Cycle},
is the compiler permitted to act as if the code was instead as shown in
Listing~\ref{lst:Unused Inventing Non-Boolean Atomic Loads}?
Here, if the value loaded by line~9 is something other than 42,
the load is retried on line~11, and perhaps a third time on line~14.
If the value loaded in all three cases is something other than 42,
then line~16 stores the last value loaded, and otherwise, the constant
42 is stored by either line~18, 20, or~23.
Despite the fact that the code in both listings can be argued to be
equivalent, it is difficult to argue that a reasonable compiler
would act as if the code in
Figure~\ref{lst:OOTA Cycle}
was instead that shown in
Listing~\ref{lst:Unused Inventing Non-Boolean Atomic Loads}.

@@

This section departs from one of the rules for this paper by considering
an example having undefined behavior.

@@

% @@@ reviewer to do.
% Luc Maranget: Memory models.
% Christoph Hellwig: Why slow on BPF memory model.
% Miguel Ojeda: LKMM and Rust.
% Alice Ryhl: LKMM and Rust.
% Ralf Jung: Rust.
% Catalin Marinas: ARM and formal methods (qspinlock).
% Segher Boessenkool:  GCC.
% Dan Lustig: NVIDIA memory ordering.
% Andrea Parri: Memory models, RISC-V hardware.
% Jonas Oberhauser: Memory models.
% Hernan Ponce de Leon: Memory models.
% Palmer Dabbelt: RISC-V hardware.
% Ali Sezgin: P0422R0.
% Tony Tye: P0422R0.
% Other memory-model maintainers.
% Derek Williams: PowerPC hardware.
% Hans Boehm: Skeptic, memory models, OOTA.  Suggestions for others?
% Luke Geeson: Whatever...  Review.
% Ori Lahav: Prior OOTA.
% Viktor Vafeiadis:  Prior OOTA.
% Derek Dreyer: Prior OOTA.
% Brian Demsky: Prior OOTA.

% Done as of December 18, 2023 %% as of March 18, 2023
%% David Goldblatt
%% Boqun Feng, Uladzislau Rezki, Neeraj Upadhyay, Joel Fernandes, and
%	Frederic Weisbecker (RCU proteges)
%% Michael Wong
%% Maged Michael
% John Wickerson: Thank you.  OK for Peter Sewell, Mark Batty, etc.
%% Akira Yokosawa: Professional courtesy, perfbook memory models.
%% Dan Kelley, Alexei Starovoitov, Mykola Lysenko: FYI.  OOTA dropped
% Alan Jeffrey: Fixed-point insight P0422R0.  Reached out via LinkedIn.
%% Jade Alglave: C++ herd model?  Memory models and variety of hardware.
%% Alan Stern: Memory models, mathematical logic, and variety of hardware.
% Martin Uecker (offered)
% Peter Sewell
% Mark Batty
% Richard Gristenthwaite: ARM hardware.  Posed HW speculation question.
% Peter Zijlstra.
% Mark Rutland: ARM memory ordering.
% Will Deacon: ARM hardware, C++, and OOTA.
%% Greg Kroah-Hartman and Linus Torvalds: FYI, trouble being caused.
%% Olivier Giroux: Skeptic, memory models, NVIDIA, requested.
%% Gonzalo Brito <gonzalob@nvidia.com>
